---
title: "Severe Weather Events Can Cause Both Public Health and Economic Problems in US."
author: "Sunny Pan"
date: "08/08/2020"
output: html_document
---

## Synopsis
This project determine which weather event is the most harmful to public health and the greatest economic consequences in US. It uses the data from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

## Data Processing
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(reshape2)
```
### Loading the data
The data download from [this link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

The explanation of the data check [this link](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

```{r, cache=TRUE}
myfile <- "repdata_data_StormData.csv.bz2"
orgdata <- read.csv(myfile, header = TRUE, sep = ",", stringsAsFactors = FALSE)
```
The explanation of the data Check [this link](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

### Question1: Across the United States, which types of events are most harmful with respect to population health?
We will use the variables "EVTYPE","FATALITIES" and "INJURIES" to find out the most harmful event to population health.

1. Select the variables "EVTYPE","FATALITIES" and "INJURIES" from original data.
```{r}
mydata <- orgdata %>% 
        select("EVTYPE","FATALITIES","INJURIES") %>% 
        filter(!is.na(EVTYPE), !is.na(FATALITIES), !is.na(INJURIES), INJURIES!=0|FATALITIES!=0) %>%
        mutate(TOTALHARM = INJURIES+FATALITIES)
mysum <- mydata %>% 
        group_by(EVTYPE) %>%
        summarise(minfatal = min(FATALITIES), meanfatal = mean(FATALITIES), 
                  midianfatal = median(FATALITIES), maxfatal = max(FATALITIES),
                  mininjuries = min(INJURIES), meaninjuries = mean(INJURIES), 
                  midianjuries = median(INJURIES), maxjuries = max(INJURIES),
                  mintotal = min(TOTALHARM), meantotal = mean(TOTALHARM), 
                  midiantotal = median(TOTALHARM), maxtotal = max(TOTALHARM))
```

2. Top 10 fatalities events.
```{r}
mysum1 <- mysum %>% arrange(desc(meanfatal, meantotal, meaninjuries))
top10fatalevn <- mysum1$EVTYPE[1:10]
top10fatal <- mydata %>% 
        filter(EVTYPE %in% top10fatalevn) %>% 
        arrange(desc(FATALITIES, EVTYPE))
```
The top 10 fatalities events to the population health are `r top10fatalevn`. 

3. Top 10 injuries events.
```{r}
mysum2 <- mysum %>% arrange(desc(meaninjuries, meantotal, meanfatal))
top10injuriesevn <- mysum2$EVTYPE[1:10]
top10injuries <- mydata %>% 
        filter(EVTYPE %in% top10injuriesevn) %>% 
        arrange(desc(INJURIES, EVTYPE))
```
The top 10 injuries events to the population health are `r top10injuriesevn`. 

4. Top 10 total harmful events(fatalities + injuries).
```{r top10total}
mysum3 <- mysum %>% arrange(desc(meantotal, meanfatal, meaninjuries))
top10totalevn <- mysum3$EVTYPE[1:10]
top10total <- mydata %>% 
        filter(EVTYPE %in% top10totalevn) %>% 
        arrange(desc(TOTALHARM, EVTYPE))
```
The top 10 harmful events to the population health is `r top10totalevn`. 

### Question2: Across the United States, which types of events have the greatest economic consequences?
We will calculate the damage of property and crop using the variables "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP". "PROPDMG" and "PROPDMGEXP" corresponds to the damage in dollars, exponential factor for property damage (“K” for thousands, “M” for millions, “B” for billions), "CROPDMG", "CROPDMGEXP" corresponds to the damage in dollars, exponential factor for crop damage. 

1. Select the variables "EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP" from original data
```{r}
mydamage <- orgdata %>% 
        select("EVTYPE","PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP") %>% 
        filter(!is.na(EVTYPE), !is.na(PROPDMG), !is.na(CROPDMG), PROPDMG!=0|CROPDMG!=0)
```
2. Calculate the actual damage values
```{r}
mydamage1 <- mydamage %>% mutate(
        PROPDMG1 = case_when(PROPDMGEXP == "K"|PROPDMGEXP == "k"~PROPDMG*1000,
                             PROPDMGEXP == "M"|PROPDMGEXP == "m"~PROPDMG*1000000,
                             PROPDMGEXP == "B"|PROPDMGEXP == "k"~PROPDMG*1000000000,
                             TRUE~0),
        CROPDMG1 = case_when(CROPDMGEXP == "K"|CROPDMGEXP == "k"~CROPDMG*1000,
                             CROPDMGEXP == "M"|CROPDMGEXP == "m"~CROPDMG*1000000,
                             CROPDMGEXP == "B"|CROPDMGEXP == "k"~CROPDMG*1000000000,
                             TRUE~0)
        )
mydamage1 <- mydamage1 %>%
        mutate(TOTALDMG = PROPDMG1 + CROPDMG1)%>%
        select(EVTYPE, PROPDMG = PROPDMG1, CROPDMG = CROPDMG1, TOTALDMG)
```
3. The top 10 damage events.
```{r top10DMG, echo=TRUE}
sumdamage1 <- mydamage1 %>%
        group_by(EVTYPE) %>%
        summarise(meanprop=mean(PROPDMG),medprop=median(PROPDMG),
                  meancrop=mean(CROPDMG),medcrop=median(CROPDMG),
                  meantotal=mean(TOTALDMG),medtotal=median(TOTALDMG)
                  )
top10damageevn <- arrange(sumdamage1, desc(meantotal))$EVTYPE[1:10]
top10damage <- mydamage1 %>% 
        filter(EVTYPE %in% top10damageevn) %>% 
        arrange(desc(TOTALDMG, EVTYPE))
```
The top 10 damage events to the economy is `r top10damageevn`.        

## Results
### Result1:
1. The plot for top 10 harmful events:
```{r}
sumTotal <- top10total %>% 
        group_by(EVTYPE) %>%
        summarise(TotalHarm=sum(TOTALHARM),Fatal=sum(FATALITIES),Injuries=sum(INJURIES)) %>%
        arrange(desc(TotalHarm,Fatal,Injuries))
sumTotal1 <- melt(sumTotal, variable.name = "Sum", value.name = "Num")
ggplot(sumTotal1, aes(x = EVTYPE,y = Num,fill = Sum))+
        geom_bar(stat ="identity",width = 0.8,position = "dodge")+
        scale_fill_manual(values = c("green","red","blue"))+
        labs(x = "Top 10 Harmful Event",y = "Number of Harm", title = "Top 10 Harmful Events to Population Health")+
        geom_text(aes(label = Num),position=position_dodge(width = 0.8),size = 3,vjust = -0.25)+
        theme(text = element_text(size=10), axis.text.x = element_text(angle=90))
```

2. The plot for top 10 fatal events:
```{r}
sumTotal2 <- top10fatal %>% 
        group_by(EVTYPE) %>%
        summarise(Fatal=sum(FATALITIES),TotalHarm=sum(TOTALHARM),Injuries=sum(INJURIES)) %>%
        arrange(desc(Fatal, TotalHarm, Injuries))
sumTotal22 <- melt(sumTotal2, variable.name = "Sum", value.name = "Num")
ggplot(sumTotal22, aes(x = EVTYPE,y = Num,fill = Sum))+
        geom_bar(stat ="identity",width = 0.8,position = "dodge")+
        scale_fill_manual(values = c("red", "green", "blue"))+
        labs(x = "Top 10 Fatal Event",y = "Number of Harm", title = "Top 10 Fatal Events to Population Health")+
        geom_text(aes(label = Num),position=position_dodge(width = 0.8),size = 3,vjust = -0.25)+
        theme(text = element_text(size=10), axis.text.x = element_text(angle=90))
```

From the plots, we can see the most harmful event is "HURRICANE/TYPHOON" to population health and the "HEAT WAVE" is the most fatality event.

### Result2:
1. The plot of top 10 damage events to enconomy:
```{r}
sumTotal <- top10damage %>% 
        group_by(EVTYPE) %>%
        summarise(TOTALDMG=round(sum(TOTALDMG)/1000000000,1),
                  PROPDMG=round(sum(PROPDMG)/1000000000,1),
                  CROPDMG=round(sum(CROPDMG)/1000000000,1)) %>%
        arrange(desc(TOTALDMG))
sumTotal1 <- melt(sumTotal, variable.name = "Sum", value.name = "Num")
ggplot(sumTotal1, aes(x = EVTYPE,y = Num,fill = Sum))+
        geom_bar(stat ="identity",width = 1,position = "dodge")+
        scale_fill_manual(values = c("green","red","blue"))+
        labs(x = "Event Name",y = "Number of Damage(Billion Dollars)", title = "Top 10 Damage Events to economy")+
        geom_text(aes(label = Num),position=position_dodge(width = 1),size = 3,vjust = -0.25)+
        theme(text = element_text(size=10), axis.text.x = element_text(angle=90))
```
    
From the plot, we can see the most damage event is "HURRICANE/TYPHOON" to the economic consequences across the United States.
